---
title: "Ecotopic 2 Main Markdown"
author: "Stellaria Otstot & Michael Birch-Tanzer"
date: "2026-03-03"
output: html_document
---

### Research Question

How has lake mead’s water level decrease affected water fowl observations?

> Hypothesis 1: A decrease in the water level of lake mead will be quantitatively associated with a decrease in ebird observations of waterfowl compared to total lake mead observations because waterfowl rely on this critical watershed for habitat, food, and breeding. 

> Hypothesis 2: In a reservoir unaffected by water depletion the ratio of waterfowl observations to total observations will remain relatively constant because they don’t face this unique environmental pressure.

> Hypothesis 3: In study sites similar to the conditions found in lake mead (ie, stressed conditions), we will see a decline in the ratio of waterfowl observations to total observations.

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages
```{r}
#load in necessary packages
library(sf)
library(tmap)
library(lubridate)
library(dplyr)
library(ggplot2)

getwd()
```

### Load in Data
```{r}
# Read in bird data from Nevada
NV_bird_data <- read.csv("data/NV_final_birds.csv")

# Read in lake mead water level data
water_elevation <- read.csv("data/lake_mead_water_elevation.csv")
#head(water_elevation)
  #filter or combine average for fall/spring
water_elevation$datetime <- as.Date(water_elevation$datetime)
water_elevation$year <- year(water_elevation$datetime)
water_elevation$month<- month(water_elevation$datetime)

water_elevation_season<- water_elevation %>%
  mutate(season = case_when(
    month %in% c(2,3,4,5) ~ "Spring",
    month %in% c(8,9,10,11) ~ "Fall",
    ))




# Read in shape file for study location (around lake mead) (Q: can we download individual data points rather than shape file)
lake_mead_shape <- st_read("data/lake_tracts/lake_tracts.shp")
sf::sf_use_s2(FALSE) #solving spherical geom problem
mega_lake <- st_union(lake_mead_shape) #dissolving polygon borders into one mega polygon
mini_lakes <- st_cast(mega_lake, "POLYGON") #splits up multi-polygon into individual polygons
lake_mead <- mini_lakes[2, ] #keep specific lake mead polygon (row 2, all columns)
tm_shape(lake_mead) + tm_polygons() 
```

# Data Wrangling

### Data set Wrangling
```{r}
# Clip bird data to lake mead shape file boundary (just include bird data for study location)
# Pre visualization of lake mead shape file boundary
# Drop spatial geometry

### all done in outside rmd (read in as "NV_final_birds.csv")
```


### Variable Wrangling
```{r}
head(NV_bird_data)

#make new variable that assigns species to binary categories (waterfowl_yes, waterfowl_no)
NV_bird_data <- NV_bird_data %>%
  mutate(waterfowl = ifelse(common_name %in% c("Western Grebe", "American Coot", "Ring-billed Gull", "American White Pelican"),
      "yes",
      "no"))

#group by year and month
NV_bird_basics <- NV_bird_data %>%
  # Fix observation_count
  mutate(
    observation_count = ifelse(observation_count == "X", 1, as.numeric(observation_count)),
    observation_date = as.Date(observation_date),
    year = as.numeric(format(observation_date, "%Y")),
    month = as.numeric(format(observation_date, "%m")),
    season = case_when(
      month %in% 2:5  ~ "Spring",
      month %in% 8:11 ~ "Fall",
      TRUE             ~ NA_character_  # ignore other months
    )
  ) %>%
  filter(!is.na(season)) %>%
  group_by(common_name, year, season, waterfowl) %>%
  summarise(total_observations = sum(observation_count, na.rm = TRUE)) %>%
  ungroup()


#make a proportion of waterfowl to all birds (waterfowl_prop)
NV_bird_prop <- NV_bird_data %>%
  # Fix observation_count and extract year/season
  mutate(
    observation_count = ifelse(observation_count == "X", 1, as.numeric(observation_count)),
    observation_date = as.Date(observation_date),
    year = as.numeric(format(observation_date, "%Y")),
    month = as.numeric(format(observation_date, "%m")),
    season = case_when(
      month %in% 2:5  ~ "Spring",
      month %in% 8:11 ~ "Fall",
      TRUE             ~ NA_character_
    )
  ) %>%
  filter(!is.na(season)) %>%
  # Group by year and season only
  group_by(year, season) %>%
  # Summarize total waterfowl and non-waterfowl
  summarise(
    waterfowl_total = sum(observation_count[waterfowl == "yes"], na.rm = TRUE),
    non_waterfowl_total = sum(observation_count[waterfowl == "no"], na.rm = TRUE),
    total_birds = waterfowl_total + non_waterfowl_total,
    waterfowl_prop = waterfowl_total / total_birds
  ) %>%
  ungroup()

prop_and_water <- NV_bird_prop %>%
  left_join(water_elevation_season, by = c("year", "season"))

```

# Body

### Preliminary Visualizations
```{r}
# scatterplot: waterfowl_yes and waterfowl_no vs time (similar to below)

# scatterplot: waterfowl_prop vs time
NV_bird_prop %>%
  ggplot(aes(y = waterfowl_prop, x = year, color = season)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

# scatterplot: water_level vs time


# scatterplot: waterfowl_prop vs water_level

```


### Model Building
```{r}
#waterfowl prop vs time
model_1 <- lm(waterfowl_prop ~ year + season, data = NV_bird_prop)
summary(model_1)
```

### Next steps?
```{r}
# read in more study locations (more in az/nv)
## other reseviors (one similar with regard to water stress and one unstressed)
## evaluate water level during spring and in fall
```



```{r}
tmap_mode("view")
LakeMead <- tm_shape(lake_mead_shape) +
  tm_polygons("TRACT_ID") 

LakeMead

```