---
title: "Ecotopic 2 Main Markdown"
author: "Stellaria Otstot & Michael Birch-Tanzer"
date: "2026-03-03"
output: html_document
---

### Research Question

How has lake mead’s water level decrease affected water fowl observations?

> Hypothesis 1: A decrease in the water level of lake mead will be quantitatively associated with a decrease in ebird observations of waterfowl compared to total lake mead observations because waterfowl rely on this critical watershed for habitat, food, and breeding. 

> Hypothesis 2: In a reservoir unaffected by water depletion the ratio of waterfowl observations to total observations will remain relatively constant because they don’t face this unique environmental pressure.

> Hypothesis 3: In study sites similar to the conditions found in lake mead (ie, stressed conditions), we will see a decline in the ratio of waterfowl observations to total observations.

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages
```{r}
#load in necessary packages
library(sf)
library(tmap)
library(lubridate)
library(dplyr)
library(ggplot2)

#getwd()
```

### Load in Data
```{r}
# Read in bird data from Nevada
NV_bird_data <- read.csv("data/NV_final_birds.csv")

# Read in lake mead water level data
water_elevation <- read.csv("data/lake_mead_water_elevation.csv")
#head(water_elevation)

# Read in shape file for study location (around lake mead)
lake_mead_shape <- st_read("data/lake_tracts/lake_tracts.shp")
sf::sf_use_s2(FALSE) #solving spherical geom problem
mega_lake <- st_union(lake_mead_shape) #dissolving polygon borders into one mega polygon
mini_lakes <- st_cast(mega_lake, "POLYGON") #splits up multi-polygon into individual polygons
lake_mead <- mini_lakes[2, ] #keep specific lake mead polygon (row 2, all columns)
tm_shape(lake_mead) + tm_polygons() 
```

# Data Wrangling

### Data set Wrangling

#### Birds
```{r}
# bird data
    # Clip bird data to lake mead shape file boundary (just include bird data for study location)
    # Pre visualization of lake mead shape file boundary
    # Drop spatial geometry

    ### all done in outside rmd (read in as "NV_final_birds.csv")
```


#### Water Level
```{r}
# water elevation data
water_elevation$datetime <- as.Date(water_elevation$datetime)
water_elevation$year <- year(water_elevation$datetime)
water_elevation$month<- month(water_elevation$datetime)

water_elevation_season<- water_elevation %>%
  # filter or combine average for fall/spring
  mutate(season = case_when(
    month %in% c(2:5) ~ "Spring",
    month %in% c(8:11) ~ "Fall",
    TRUE ~ NA_character_
    ))%>%
  filter(!is.na(season))  # only keep spring/fall months

# Compute seasonal averages per year
water_elevation_mean <- water_elevation_season %>%
  group_by(year, season) %>%
  summarise(
    mean_water_level = mean(pool.elevation, na.rm = TRUE),
    .groups = "drop"
  )
```


#### Join Birds and Water
```{r}
NV_bird_data <- NV_bird_data %>%
  mutate(waterfowl = ifelse(common_name %in% c("Western Grebe", "American Coot", "Ring-billed Gull", "American White Pelican"),
      "yes",
      "no"))

final_data <- NV_bird_data %>%
  mutate(
    #fix obs with X's
    observation_count = ifelse(observation_count == "X", 1, as.numeric(observation_count)), 
    #lubridate breaks apart date column
    observation_date = as.Date(observation_date),
    year = as.numeric(format(observation_date, "%Y")),
    month = as.numeric(format(observation_date, "%m")), 
    #make categories for spring and fall
    season = case_when(
      month %in% 2:5  ~ "Spring",
      month %in% 8:11 ~ "Fall", 
      #gets rid of observations that do not fall into these spring/fall categories
      TRUE ~ NA_character_ 
    )
  ) %>%
  #filter out na's
  filter(!is.na(season)) %>%
  #group by and summarize vars of interest
  group_by(common_name, year, season, waterfowl) %>%
  summarise(
    #make obs count
    total_observations = sum(observation_count, na.rm = TRUE),
    #drop groups
    .groups = "drop" 
  ) %>%
  group_by(year, season) %>%
  summarise(
    #make vars for prop calculations
    waterfowl_total = sum(total_observations[waterfowl == "yes"], na.rm = TRUE),
    non_waterfowl_total = sum(total_observations[waterfowl == "no"], na.rm = TRUE),
    total_birds = waterfowl_total + non_waterfowl_total,
    waterfowl_prop = waterfowl_total / total_birds,
    .groups = "drop"
  ) %>%
  #join water elevation data by year and season
  left_join(water_elevation_mean, by = c("year", "season"))

```

# Body

### Preliminary Visualizations
```{r}
# scatterplot: waterfowl and non_waterfowl vs time (similar to below)
final_data %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = waterfowl_total, color = "Waterfowl")) +
  geom_point(aes(y = non_waterfowl_total, color = "Non-waterfowl")) +
  geom_smooth(aes(y = waterfowl_total, color = "Waterfowl"), method = "lm") +
  geom_smooth(aes(y = non_waterfowl_total, color = "Non-waterfowl"), method = "lm") +
  theme_minimal()

# scatterplot: waterfowl_prop vs time
final_data %>%
  ggplot(aes(y = waterfowl_prop, x = year, color = season)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Proportion of Waterfowl Observations to Non-waterfowl Observations Over Time") +
  ylab("waterfowl to non-waterfowl proportion")


# scatterplot: water_level vs time
final_data %>%
  ggplot(aes(y = mean_water_level, x = year, color = season))+
  geom_point() +
  geom_smooth(se = FALSE, alpha = 0.5, #color = "black")
  ) +
  theme_minimal()+
  labs(title = "Lake Mead Water Level Over Time") +
  ylab("water level (feet)")

# scatterplot: waterfowl_prop vs water_level
final_data %>%
  ggplot(aes(y = waterfowl_prop, x = mean_water_level, color = season)) +
  scale_x_reverse()+
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()+
  labs(title = "Proportion of Waterfowl to Non-waterfowl Observations vs Water Level") +
  ylab("waterfowl to non-waterfowl proportion") +
  xlab("water level (feet)")

```


### Model Building
```{r}
#waterfowl prop vs time
model_1 <- lm(waterfowl_prop ~ year + season, data = NV_bird_prop)
summary(model_1)

#waterfowl prop vs pool elevation
model_2 <- lm(waterfowl_prop ~ pool.elevation + season, data = prop_and_water)
summary(model_2)
```

### Next steps?
```{r}
# read in more study locations (more in az/nv)
## other reseviors (one similar with regard to water stress and one unstressed)
## evaluate water level during spring and in fall
```


# Apendix

### lake mead shape
```{r}
tmap_mode("plot")
LakeMead <- tm_shape(lake_mead_shape) +
  tm_polygons("TRACT_ID") 

LakeMead
```