---
author: "Stellaria Otstot & Michael  Birch-Tanzer"
date: "2026-03-03"
output: html_document
---

# Introduction

### Research Question

How has lake mead’s water level decrease affected water fowl observations?

> Hypothesis 1: A decrease in the water level of lake mead will be quantitatively associated with a decrease in ebird observations of waterfowl compared to total lake mead observations because waterfowl rely on this critical watershed for habitat, food, and breeding. 

> Hypothesis 2: In a reservoir unaffected by water depletion the ratio of waterfowl observations to total observations will remain relatively constant because they don’t face this unique environmental pressure.

> Hypothesis 3: In study sites similar to the conditions found in lake mead (ie, stressed conditions), we will see a decline in the ratio of waterfowl observations to total observations.

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages
```{r}
#load in necessary packages

getwd()
```

### Load in Data
```{r}
# Read in bird data from Nevada and Arizona 

# Read in lake mead water level data
water_elevation <- read.csv("lake_mead_water_elevation.csv")
head(water_elevation)
  #filter or combine average for fall/spring

# Read in shape file for study location (around lake mead) (Q: can we download individual data points rather than shape file)
lake_mead_shape <- st_read("lake_tracts/lake_tracts.shp")
sf::sf_use_s2(FALSE) #solving spherical geom problem
mega_lake <- st_union(lake_mead_shape) #dissolving polygon borders into one mega polygon
mini_lakes <- st_cast(mega_lake, "POLYGON") #splits up multi-polygon into individual polygons
lake_mead <- mini_lakes[2, ] #keep specific lake mead polygon (row 2, all columns)
tm_shape(lake_mead) + tm_polygons() 
```

# Data Wrangling

### Data set Wrangling
```{r}
# Clip bird data to lake mead shape file boundary (just include bird data for study location)

# Pre visualization of lake mead shape file boundary

# Drop spatial geometry
## st_drop_geometry()
```

### Variable Wrangling
```{r}
#make new variable that assigns species to binary categories (waterfowl_yes, waterfowl_no)

#make a proportion of waterfowl to all birds (waterfowl_prop)
```

# Body

### Preliminary Visualizations
```{r}
# scatterplot: waterfowl_yes and water_fowl vs time (similar to below)

# scatterplot: waterfowl_prop vs time

# scatterplot: water_level vs time

# scatterplot: waterfowl_prop vs water_level
```


### Model Building
```{r}
# model <- lm(y ~ x)
```

### Next steps?
```{r}
# read in more study locations (more in az/nv)
## other reseviors (one similar with regard to water stress and one unstressed)
## evaluate water level during spring and in fall
```



```{r}
tmap_mode("view")
LakeMead <- tm_shape(lake_mead_shape) +
  tm_polygons("TRACT_ID") 

LakeMead

```